---
title: "GPvecchia tutorial"
author: "Marcin Jurek, Daniel Zilber, Matthias Katzfuss"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GPvecchia}
  % \VignetteDepends{mvtnorm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.dim=c(7,5)
)
```

In this vignette, we will demonstrate the main capabilities of the `GPvecchia` package. These include estimating parameters and making spatial predictions. We also show how to use the package for processing non-linear, non-Gaussian data by combining the Vecchia with a Laplace approximation.

We start by importing the `GPvecchia` library.
```{r}
library(GPvecchia)
library(Matrix)
library(fields)
```


### Simulating data for illustration

To illustrate our method, we simulate a small dataset. First, consider a unit square and randomly select observation locations. (Set spatial.dim=1 to consider data on the one-dimensional unit interval.)
```{r}
set.seed(1988)
spatial.dim=2
n=50
if(spatial.dim==1){
  locs=matrix(runif(n),ncol=1)
} else {
  locs <- cbind(runif(n),runif(n))
}
```

Next, we define the covariance function of the field as well as the scale of the measurement error
```{r}
beta=2
sig2=1; range=.1; smooth=1.5
covparms =c(sig2,range,smooth)
covfun <- function(locs) sig2*MaternFun(fields::rdist(locs),covparms)
nuggets=rep(.1,n)
```

We are now ready to simulate the field and visualize it as a sanity check.
```{r fig4, out.width = '400px'}
Om0 <- covfun(locs)+diag(nuggets)
z=as.numeric(t(chol(Om0))%*%rnorm(n))
data=z+beta

# plot simulated data
if(spatial.dim==1) {
  plot(locs,data)
} else {
  fields::quilt.plot(locs,data)
}
```

We also create a grid of $n.p$ locations at which we would like to make predictions.
```{r}
n.p=100
if(spatial.dim==1){  #  1-D case
  locs.pred=matrix(seq(0,1,length=n.p),ncol=1)
} else {   # 2-D case
  grid.oneside=seq(0,1,length=round(sqrt(n.p)))
  locs.pred=as.matrix(expand.grid(grid.oneside,grid.oneside)) # grid of pred.locs
}
n.p=nrow(locs.pred)
```


### Basic functions for parameter estimation and prediction

Let us now estimate the mean and covariance parameters using the default settings, which assume a spatially constant mean or trend, and a Matern covariance structure. Note that the following code might take a minute or so to run.
```{r paramm-est, cache=TRUE}
vecchia.est=vecchia_estimate(data,locs)
```
Based on these parameter estimates, we can then make predictions at the grid of locations we had specified above.
```{r}
preds=vecchia_pred(vecchia.est,locs.pred)
```

Finally, we compare the approximate predictions with the best possible ones (i.e. those obtained using analytic expressions for conditional mean in the Gaussian distribution). 
```{r}
##  exact prediction
mu.exact=as.numeric(covfun(rbind(locs,locs.pred))[,1:n]%*%solve(Om0,data-beta))+beta
cov.exact=covfun(rbind(locs,locs.pred))-
  covfun(rbind(locs,locs.pred))[,1:n]%*%solve(Om0,t(covfun(rbind(locs,locs.pred))[,1:n]))
var.exact=diag(cov.exact)
cov.exact.pred=cov.exact[n+(1:n.p),n+(1:n.p)]


### plot Vecchia and exact predictions
if(spatial.dim==1) {
  plot(locs,z)
  lines(locs.pred,preds$mean.pred,col='blue')
  lines(locs.pred,preds$mean.pred-1.96*sqrt(preds$var.pred),col='blue',lty=2)
  lines(locs.pred,preds$mean.pred+1.96*sqrt(preds$var.pred),col='blue',lty=2)
  lines(locs.pred,mu.exact[n+(1:n.p)],col='red')
  lines(locs.pred,mu.exact[n+(1:n.p)]-1.96*sqrt(var.exact[n+(1:n.p)]),col='red',lty=2)
  lines(locs.pred,mu.exact[n+(1:n.p)]+1.96*sqrt(var.exact[n+(1:n.p)]),col='red',lty=2)
} else {
  sdrange=range(sqrt(c(preds$var.pred,var.exact[n+(1:n.p)])))
  defpar = par(mfrow=c(2,3))
  fields::quilt.plot(locs,z)
  fields::quilt.plot(locs.pred,preds$mean.pred)
  fields::quilt.plot(locs.pred,sqrt(preds$var.pred),zlim=sdrange)
  fields::quilt.plot(locs,z)
  fields::quilt.plot(locs.pred,mu.exact[n+(1:n.p)])
  fields::quilt.plot(locs.pred,sqrt(var.exact[n+(1:n.p)]),zlim=sdrange)
  par(defpar)
}
```


### More details on likelihood evaluation

Let's take a closer look at how the likelihood is evaluated using Vecchia. Most importantly, we can specify a parameter, $m$. Its value determines the number of "neighbours" of each point, or, in other words, how many other points a given point conditions on. The larger this parameter, the more accurate and expensive the approximation will be.
```{r}
m=20
vecchia.approx=vecchia_specify(locs,m)
vecchia_likelihood(z,vecchia.approx,covparms,nuggets)
```
Note that the function vecchia_specify determines the general properties of the approximation, but it does not depend on the data or the specific parameter values. Hence, it does not have to be re-run when searching over different parameter values in an estimation procedure.

We can also compare the results to the exact likelihood:
```{r}
library(mvtnorm)
dmvnorm(z,mean=rep(0,n),sigma=Om0,log=TRUE)
```
In this case the approximation is very good. In general, $m=20$ is a good value, and $m$ should usually be between 10 and 40. For one-dimensional space, we can get good approximations even with $m=5$ or smaller.


### More details on spatial prediction

Similar to the previous section we next specify the approximation and indicate at which locations prediction is desired.
```{r}
m=30
vecchia.approx=vecchia_specify(locs,m,locs.pred=locs.pred)
preds=vecchia_prediction(z,vecchia.approx,covparms,nuggets)
# returns a list with elements mu.pred,mu.obs,var.pred,var.obs,V.ord
```

It is also possible to print the entire predictive covariance matrix. We do it here only for the purpose of illustration. If $n.p$ is very large, this matrix might use up a lot of memory and we generally do not recommend plotting it directly.
```{r}
Sigma=V2covmat(preds)$Sigma.pred
cov.range=quantile(rbind(Sigma,cov.exact.pred),c(.01,.99))
defpar = par(mfrow=c(1,2))
fields::image.plot(cov.exact.pred,zlim=cov.range)
fields::image.plot(Sigma,zlim=cov.range)
par(mfrow=c(defpar))
```

#### Linear combinations

We might sometimes be interested in a linear combination of the predicted values. In particular, we can limit our attention to only a subset of our predictions. This can be accomplished by specifying the linear combination coefficients as a matrix. As an example, we assume we are only interested in predictions at the unobserved prediction locations (not at the first n observed locations):
```{r}
H=Matrix::sparseMatrix(i=1:(n+n.p),j=1:(n+n.p),x=1)[(n+1):(n+n.p),]

# compute variances of Hy
lincomb.vars=vecchia_lincomb(H,preds$U.obj,preds$V.ord)
plot(preds$var.pred,lincomb.vars)
```

As another example, we consider the overall mean of the process at all prediction locations. Using the `vecchia_lincomb()` function enables us to get the variance estimates easily. 
```{r}
mean(preds$mu.pred)

# compute entire covariance matrix of Hy (here, 1x1)
H=Matrix::sparseMatrix(i=rep(1,n.p),j=n+(1:n.p),x=1/n.p)
lincomb.cov=vecchia_lincomb(H,preds$U.obj,preds$V.ord,cov.mat=TRUE)
```



### Other GP approximations as special cases

By specifying appropriate options in vecchia_specify, we can do everything described above for several other GP approximations: Modified predictive process, FSA, MRA, latent, standard Vecchia

Setting $M=1$ results in **block full-scale approximation**, specifically one with $r_0 = \frac{m}{2}$ knots spread over the entire domain and the remaining locations being partitioned into blocks of size $<\frac{m}{2}+1$.
```{r}
m=20
mra.options.fulls=list(M=1)
blockFS = vecchia_specify(locs, m, 'maxmin', conditioning='mra', mra.options=mra.options.fulls, verbose=TRUE)
```

Another popular existing approximation method, **modified predictive process (MPP)**, can also be obtained by specifying appropriate parameter settings:
```{r}
mra.options.mpproc=list(r=c(m,1))
MPP = vecchia_specify(locs, m, 'maxmin', conditioning='mra', mra.options=mra.options.mpproc, verbose=TRUE)
```
As we can see, MPP can be viewed as a special case of the **multi-resolution approximation (MRA)**.

A general MRA is obtained my specifying all of its three parameters
```{r}
mra.options.mra = list(r=c(10, 5, 5), M=2, J=2)
MRA_rJM = vecchia_specify(locs, m, 'maxmin', conditioning='mra', mra.options=mra.options.mra, verbose=TRUE)
```
We should note two things to note about this full specifiction of an MRA. First, providing all three $r$,$J$ and $M$ overrides whatever value of $m$ was provided. Second, in order to be able to place a knot at each point of the grid, the provided parameters might need to be adjusted. 

Finally, we can also use  the `GPvecchia` package to specify a **Nearest Neighbour Gaussian Process (NNGP)** approximation. This can be accomplished as shown below. 
```{r}
NNGP = vecchia_specify(locs, m, cond.yz='y')
```


We can now easily compare different approximation methods and compare it with SGV and exact likelihood.
```{r}
vecchia_likelihood(z,blockFS,covparms,nuggets)
vecchia_likelihood(z,MPP,covparms,nuggets)
vecchia_likelihood(z,MRA_rJM,covparms,nuggets)
vecchia_likelihood(z,NNGP,covparms,nuggets)
vecchia_likelihood(z, vecchia_specify(locs, m), covparms, nuggets)
dmvnorm(z,mean=rep(0,n),sigma=Om0,log=TRUE)
```



### Non-Gaussian data
Here we demonstrate how GPVecchia can fit a latent model to non-Gaussian data using the Vecchia-Laplace method.  We simulate data by first generating a correlated latent field without noise, assuming the same covariance and locations generated earlier:

```{r}
# simulate latent process
y=as.numeric(t(chol(Om0))%*%rnorm(n))
```


Then we sample a single non-Gaussian value for each latent value.  The variability introduced by the sampling induces heteroskedasticity, in contrast the the constant noise added to the Gaussian case.  Below we use a logistic model for binary data, but there are implementations for count and continuous positive (right-skewed) data as well.

```{r}
data.model = "logistic"

# simulate data
if(data.model=='poisson'){
  z = rpois(n, exp(y))
} else if(data.model=='logistic'){
  z = rbinom(n,1,prob = exp(y)/(1+exp(y)))
} else if(data.model=='gamma'){
  z = rgamma(n, shape = default_lh_params$alpha, rate = default_lh_params$alpha*exp(-y))
}else{
  print('Error: Distribution not implemented yet.')
}

# plot simulated data, 1 or 2D
defpar = par(mfrow=c(1,2))
if(spatial.dim==1) {
  plot(locs,y, main = "latent")
  plot(locs,z, main = "observed")
} else {
  fields::quilt.plot(locs,y, main = "Latent")
  fields::quilt.plot(locs,z, main = "Observed")
}
par(defpar)

```

Given the simulated data, we now can efficiently estimate the latent field by specifying the number of conditioning points $m$ described earlier.  Interweaved ordering is best for 1D data while response-first ('zy') ordering is best for higher dimensions.  
```{r}
m=10
if(spatial.dim==1){
  vecchia.approx=vecchia_specify(locs,m) #IW ordering
} else {
  vecchia.approx=vecchia_specify(locs,m,cond.yz='zy') #RF ordering
}
```

With the approximated covariance structure, we can calculate the posterior estimate for the latent field using the Vecchia-Laplace method and plot the result.  Pure Laplace approximation is included for comparison; even with a samll value for $m$, we can get a result similar to Laplace but with much lower cost.

```{r}
posterior = calculate_posterior_VL(z,vecchia.approx,likelihood_model=data.model,
                                   covparms = covparms)
if (spatial.dim==1){
  par(mfrow=c(1,1))
  ord = order(locs) # order so that lines appear correctly
  y_limits = c(min(y, posterior$mean[ord]), max(y, posterior$mean[ord]))
  plot(locs[ord], y[ord], type = "l", ylim = y_limits )
  lines(locs[ord], posterior$mean[ord], type = "l", col=3, lwd=3)
  legend("bottomright", legend = c("Latent", "VL"), col= c(1,3), lwd=c(1,3))
} else if (spatial.dim==2){
  dfpar = par(mfrow=c(1,2))
  # ordering unnecessary; we are using a scatter plot rather than lines
  quilt.plot(locs, y, main= "Truth")
  quilt.plot(locs, posterior$mean,  main= "VL m=10")
  par(defpar)
  
}

```


### Non-Gaussian predictions

Predictions are computed as before, using Vecchia-Laplace methods where needed

```{r}
######  specify prediction locations   #######
n.p=30^2
if(spatial.dim==1){  #  1-D case
  locs.pred=matrix(seq(0,1,length=n.p),ncol=1)
} else {   # 2-D case
  grid.oneside=seq(0,1,length=round(sqrt(n.p)))
  locs.pred=as.matrix(expand.grid(grid.oneside,grid.oneside)) # grid of pred.locs
}
n.p=nrow(locs.pred)

######  specify Vecchia approximation   #######
vecchia.approx.pred = vecchia_specify(locs, m=10, locs.pred=locs.pred)
###  carry out prediction
preds = vecchia_laplace_prediction(posterior, vecchia.approx.pred, covparms)

# plotting predicitions
if (spatial.dim==1){
  defpar = par(mfrow=c(1,1))
  ord = order(locs) # order so that lines appear correctly
  plot(locs[ord], y[ord], type = "l", xlim=c(0,1.2), ylim = c(-1,3))
  lines(locs, posterior$mean, type = "p", col=4, lwd=3, lty=1)
  lines(locs.pred, preds$mu.pred, type = "l", col=3, lwd=3, lty=1)
  lines(locs.pred,preds$mu.pred+sqrt(preds$var.pred), type = "l", lty = 3, col=3)
  lines(locs.pred,preds$mu.pred-sqrt(preds$var.pred), type = "l", lty = 3, col=3)
  legend("topleft", legend = c("Latent", "VL: Pred", "VL: 1 stdev"), 
         col= c(1,3,3), lwd=c(1,2,1), lty = c(1,1,3))
  par(defpar)
} else if (spatial.dim==2){
  defpar =  par(mfrow=c(1,2))
  # ordering unnecessary; we are using a scatter plot rather than lines
  quilt.plot(locs, y, main= "True Latent", 
             xlim = c(0,1), ylim = c(0,1), nx=64, ny=64)
  quilt.plot(locs.pred, preds$mu.pred,  main= "VL Prediction",nx = 30, ny=30)
  par(defpar)
}
```


### Parameter estimation
The likelihood of the data for a set of parameters can be computed efficiently using the command below.
```{r}
vecchia_laplace_likelihood(z,vecchia.approx,likelihood_model=data.model,covparms = covparms)
```

This can be used for parameter estimation by evaluating the likelihood over a grid of parameter values or in an iterative optimization method such as Nelder-Mead.  Reparameterizing the parameters improves performance.

```{r, eval = FALSE}
# currently set up for covariance estimation 
vecchia.approx=vecchia_specify(locs, m=10, cond.yz = "zy") # for posterior
vecchia.approx.IW = vecchia_specify(locs, m=10) # for integrated likelihood
if (spatial.dim==1) vecchia.approx=vecchia.approx.IW

vl_likelihood = function(x0){
  theta = exp(x0)
  covparms=c(theta[1], theta[2], theta[3]) # sigma range smoothness
  prior_mean = 0 # can be a parameter as well
  # Perform inference on latent mean with Vecchia Laplace approximation
  vll = vecchia_laplace_likelihood(z,vecchia.approx, likelihood_model=data.model,
                                   covparms, return_all = FALSE,
                                   likparms = default_lh_params, prior_mean = prior_mean,
                                   vecchia.approx.IW = vecchia.approx.IW)
  return(-vll)

}
x0 = log(c(.07,1.88, 1.9))
vl_likelihood(x0)
# Issues with R aborting, maxit set to 1
res = optim(x0, vl_likelihood, method = "Nelder-Mead", control = list("trace" = 1, "maxit" = 1))
exp(res$par[1:3])
vl_likelihood(x0)
```